Large squeezing in Bose-Einstein condensates with large numbers of atoms

### 2013-07-17 large_number_squeezing_with_atoms.tex
Second round of proofreading tweaks

Ultracold atomic interferometers are also reaching state of the art sensitivities in precision measurements \cite{robins:2013}. While all current atom interferometry is limited by technical noise sources, it appears that BEC-based coherent sources offer the best chance of overcoming some of these issues, for the same reasons that optical lasers are typically preferred for optical interferometry \cite{szigeti:2012}. If these technical noise sources can be controlled, atom interferometers will be limited by atomic shot noise, which leads to a sensitivity that scales as the square root of the atomic flux.   From that point, sensitivity will only be improved by increasing the atomic flux, and/or using non-classical quantum states such as squeezed states to go below the shot noise limit \cite{robins:2013}.  Producing squeezed states for metrological purposes is therefore only relevant when it can be achieved in a context of large numbers of atoms.  This paper examines the generation of atomic squeezing using the inherent Kerr-like nonlinearities of ultracold atoms, and demonstrates a trap geometry in which metrologically relevant levels of squeezing are present in trap geometries that are compatible with large atomic number.

@@

\section{Model and two mode solutions}
\label{secTwoModeAnalytic}
Our goal is to describe the number squeezing possibilities of a BEC comprising two relevant internal states denoted by $|a\rangle$ and $|b\rangle$.  The Hamiltonian for a pair of coupled atomic internal modes is the spatial integral of the Hamiltonian density:
\begin{equation}
\hat{\mathcal{H}} = \sum_{j\in \{a,b\}} \hat{\psi}_j^{\dagger}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_j\right)\hat{\psi}_j 
          + \sum_{i,j}\frac{U_{i j}}{2} \hat{\psi}_i^{\dagger} \hat{\psi}_j^{\dagger} \hat{\psi}_j \hat{\psi}_i
          + \kappa \hat{\psi}_a^{\dagger} \hat{\psi}_b + \kappa^* \hat{\psi}_b^{\dagger}  \hat{\psi}_a,
\label{eqFieldHamiltonian}
\end{equation}
where $\hat{\psi}_j(\mathbf{r})$ is the annihilation field operator for a particle at position $\mathbf{r}$ and in internal state $|j\rangle$, $m$ is the mass of the atoms, $V_j(\mathbf{r})$ is the trapping potential for atoms in internal state $|j\rangle$, $U_{ij}$ describe the various inter- and intra-state nonlinearities, and $\kappa$ describes the coupling between the two states.

@@

These stochastic equations take the form
\begin{eqnarray}
\frac{d \phi_a}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_a(\mathbf{r}) + (U_{a a} n_a + U_{a b} \tilde{n}_b  \right)\phi_a   + \kappa \phi_b  \nonumber \\
\frac{d \phi_b}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_b(\mathbf{r}) + U_{b b} n_b + U_{a b} \tilde{n}_a   \right)\phi_b   + \kappa^* \phi_a
  \label{eqStochasticEquations}
\end{eqnarray}
where $n_j=\left|\phi_{j}\right|^2 -\frac{1}{dV}$, $\tilde{n}_j=\left|\phi_{j}\right|^2 -\frac{1}{2 dV}$, $dV$ is the volume element of the grid on which the simulation is carried out, and the terms proportional to $1/dV$ correspond to vacuum corrections. It is a peculiarity of the Wigner representation that, up to the vacuum correction terms, Eqs.~(\ref{eqStochasticEquations}) look identical to those of the coupled semi-classical nonlinear Schr{\"{o}}dinger equation. The quantum statistical nature of the Wigner equations enters due to the fact that Eqs.~(\ref{eqStochasticEquations}) are run many times and stochastically averaged, with each run using different set of random initial conditions describing (in this case) the noise on a coherent state.

@@

\begin{figure}
    \centering
    \includegraphics[width=10cm]{figures/Bog_validation}
\caption{Plot of the non-condensed fraction $N_{\mathbf{k}\neq0}/N$ as a function of the hold time $\tau_\text{hold}$.  The blue dashed line is the Bogoliubov integral approximation given in Eq.~\ref{eqBogFraction1D}, the purple dot-dashed line is the Bogoliubov summation expression given in Eq.~\ref{eqCondensateDepletion}, and the solid black line is the result of a truncated Wigner simulation.  Parameters: $L=600\,\mu\text{m}$, $N=4 \times 10^5$, $n_a=n_b=2\times 10^5$, $\chi_{aa} = 2.67 \times 10^{-2}\, \text{s}^{-1}$, $\chi_{ab}=\chi_{bb}=0$, $\theta=0$.}
  \label{figBogSqueezingValidation}
\end{figure}

@@

We wish to examine the scaling of the non-condensed fraction $N_{\mathbf{k}\neq0}/N$ with the physical size of the condensate.  Other system parameters such as the total number $N$ and the occupation of the $\ket{a}$ and $\ket{b}$ states at $t_1$ ($n_a$ and $n_b$, respectively) are kept constant and chosen to maximise the squeezing of the equivalent two-mode model presented in Section~\ref{secTwoModeAnalytic}.  In particular this fixes the values of $\theta$, $\phi$ and $\lambda_{ij} = \chi_{ij}\tau_\text{hold}$.  As the size of the condensate is scaled, $\hbar \chi_{ij} = U^\text{1D}_{ij}/L = U^\text{2D}_{ij}/A = U^\text{3D}_{ij}/V$ will change, but $\tau_\text{hold}$ will vary to keep $\lambda_{ij}$ constant.  Therefore $\Lambda_a = n_a\chi_{aa}\tau_\text{hold}$ will also be constant.  With this in mind, we can write $\tau_\text{hold}$ in terms of the system size and the nonlinearity $U_{aa}$.  The non-condensed fraction can then be shown to scale with the physical size of the condensate as:
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \sqrt{L} \cdot \frac{1}{N} \sqrt{\frac{M U^\text{1D}_{aa} n_a}{2 \hbar^2 \pi^2}} \Lambda_a^{3/2} \cos^2\theta f^\text{1D}(\Lambda_a) &&\propto \sqrt{L}, \\
  \frac{N^\text{2D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx 1 \cdot \frac{1}{N} \frac{M U^\text{2D}_{aa} n_a }{4 \hbar^2} \Lambda_a \cos^2\theta f^\text{2D}(\Lambda_a) &&\propto \text{constant}, \\
  \frac{N^\text{3D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \frac{1}{\sqrt{V}} \cdot \frac{1}{N} \left(\frac{M U^\text{3D}_{aa} n_a}{\hbar^2}\right)^{3/2} \frac{\Lambda_a^{1/2} \cos^2\theta f^\text{3D}(\Lambda_a)}{\sqrt{2} \pi^2} &&\propto \frac{1}{\sqrt{V}}.
\end{align}
The non-condensed fraction $N_{\mathbf{k}\neq 0}/N$ thus scales differently with trap size in each of 1D, 2D and 3D.  For one-dimensional traps, the non-condensed fraction increases with system size, whereas for 2D traps it is constant.  For three-dimensional traps, the non-condensed fraction \emph{decreases} with volume.  This suggests that the best system to use will be a three-dimensional box potential, and the system size can be increased until the condensate depletion becomes negligibly small.

Figures~\ref{fig1D_squeezing_and_bog_scaling}--\ref{fig3D_Bog_mode_occupation_scaling} show the non-condensed fraction predicted by the Bogoliubov analysis compared to that predicted by a full numerical simulation in one, two and three dimensions, as a function of the trap size. Also shown in the 1D and 2D cases is the maximum amount of number squeezing obtainable, as predicted by the numerical simulation. The full squeezing scaling curve is not shown in 3D, as for the smaller trap sizes the TWA method breaks down, and results are unreliable; for the larger trap sizes with negligible non-condensate fractions, however, the TWA simulation was possible in the large trap volume limit, where the non-condensed fraction was low. The traps are assumed to have a constant potential, which minimises the effects of the multimode issues described in section \ref{sec:MMdescription}.  We see that the squeezing does indeed degrade at large trap sizes in 1D, but remains constant in 2D. For small trap sizes, the squeezing in 1D asymptotically approaches that of the ideal 0D system.

@@

The simulation shown in Fig.~\ref{fig3D_Bog_mode_occupation_scaling} verified that the number of accessible states for three-dimensional traps decreases with volume in 3D.  This is because the energy of the ground state of the BEC is reducing faster than the density of states at low energies is increasing.  This scaling of the estimated Bogoliubov fraction in 3D suggests that strong squeezing may actually be found in very large, weakly trapped condensates, provided the trapping potentials are box-like. Such box-like traps have already been demonstrated and shown to be capable of producing BECs \cite{gaunt2013}.

@@

As a demonstration of the plausibility of such a scheme, we carried out a full 3D simulation using the geometry of the optical box potential described in Ref.~\cite{gaunt2013}, using the same potentials (a cylindrical ``box'' with length 63\,$\mu$m and diameter 30\,$\mu$m) and condensate number ($N=10^5$ $^{87}$Rb atoms). We set one scattering length to $5$\,nm and the others to zero: $a_{00}=5$\,nm, $a_{10}=a_{11}=0$. This system produced number squeezing of 15dB, and an extremely low non-condensed fraction of $4 \times 10^{-4}$. Both of these numbers are in agreement with the Bogoliubov calculations.

@@

The large Kerr-type nonlinearities present between neutral atoms in BEC present the possibility of significant squeezing, but they also cause potentially complicated multimode behaviour when large BECs are used.  Furthermore, the nonlinearities between multiple internal atomic states are not yet known for many atomic species, and often have sensitive dependence on external magnetic fields due to Feshbach resonances \cite{cornish:2000,altin:2010}.  This paper analyses the requirements for producing useful squeezing for systems with large numbers of atoms.

We have presented an analytic solution to the two-mode model of atomic fields with a strong Kerr non-linearity, allowing for a rapid estimate of the atomic number difference squeezing that can be obtained from these systems as scattering length data is obtained. In general, it predicts arbitrarily good squeezing in the two-mode limit. We then examined trap geometries in which this limit can be approached.

@@

\bibitem {roosET2006} Roos C F, Chwalla M, Kim K, Riebe M, and Blatt R Designer atoms for
quantum metrology 2006 \emph{Nature} {\bf 443}, 316

@@

>>>>

Ultracold atomic interferometers are also reaching state of the art sensitivities in precision measurements \cite{robins:2013, altinET2013}. While all current atom interferometry is limited by technical noise sources, it appears that BEC-based coherent sources offer the best chance of overcoming some of these issues, for the same reasons that optical lasers are typically preferred for optical interferometry \cite{szigeti:2012}. If these technical noise sources can be controlled, atom interferometers will be limited by atomic shot noise, which leads to a sensitivity that scales as the square root of the atomic flux.   From that point, sensitivity will only be improved by increasing the atomic flux, and/or using non-classical quantum states such as squeezed states to go below the shot noise limit \cite{robins:2013}.  Producing squeezed states for metrological purposes is therefore only relevant when it can be achieved in a context of large numbers of atoms.  This paper examines the generation of atomic squeezing using the inherent Kerr-like nonlinearities of ultracold atoms, and demonstrates a trap geometry in which metrologically relevant levels of squeezing are present in trap geometries that are compatible with large atomic number.

@@

\section{Model and two mode solutions}
\label{secTwoModeAnalytic}
Our goal is to describe the number squeezing possibilities of a BEC comprising two relevant internal states denoted by $|a\rangle$ and $|b\rangle$.  The Hamiltonian for a pair of coupled atomic internal modes is the spatial integral of the Hamiltonian density:
\begin{equation}
\hat{\mathcal{H}} = \sum_{j\in \{a,b\}} \hat{\psi}_j^{\dagger}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_j\right)\hat{\psi}_j 
          + \sum_{i,j}\frac{U_{i j}}{2} \hat{\psi}_i^{\dagger} \hat{\psi}_j^{\dagger} \hat{\psi}_j \hat{\psi}_i
          + \kappa \hat{\psi}_a^{\dagger} \hat{\psi}_b + \kappa^* \hat{\psi}_b^{\dagger}  \hat{\psi}_a,
\label{eqFieldHamiltonian}
\end{equation}
where $\hat{\psi}_j(\mathbf{r})$ is the annihilation field operator for a particle at position $\mathbf{r}$ and in internal state $|j\rangle$, $m$ is the mass of the atoms, $V_j(\mathbf{r})$ is the trapping potential for atoms in internal state $|j\rangle$, $\kappa$ describes the coupling between the two states, and the $U_{ij}$ describe the various inter- and intra-state nonlinearities and are given by
\begin{equation}
U_{ij} = \frac{4 \pi \hbar^2 m} {a_{ij}}
\label{eqUij}
\end{equation}
where $a_{ij}$ are the inter- and intra-species s-wave scattering lengths.

@@

These stochastic equations take the form
\begin{eqnarray}
\frac{d \phi_a}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_a(\mathbf{r}) + U_{a a} n_a + U_{a b} \tilde{n}_b  \right)\phi_a   + \kappa \phi_b  \nonumber \\
\frac{d \phi_b}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_b(\mathbf{r}) + U_{b b} n_b + U_{a b} \tilde{n}_a   \right)\phi_b   + \kappa^* \phi_a
  \label{eqStochasticEquations}
\end{eqnarray}
where $n_j=\left|\phi_{j}\right|^2 -\frac{1}{dV}$, $\tilde{n}_j=\left|\phi_{j}\right|^2 -\frac{1}{2 dV}$, $dV$ is the volume element of the grid on which the simulation is carried out, and the terms proportional to $1/dV$ correspond to vacuum corrections. It is a peculiarity of the Wigner representation that, up to the vacuum correction terms, Eqs.~(\ref{eqStochasticEquations}) look identical to those of the coupled semi-classical nonlinear Schr{\"{o}}dinger equation. The quantum statistical nature of the Wigner equations enters due to the fact that Eqs.~(\ref{eqStochasticEquations}) are run many times and stochastically averaged, with each run using different set of random initial conditions describing (in this case) the noise on a coherent state.

@@

\begin{figure}
    \centering
    \includegraphics[width=10cm]{figures/Bog_validation}
\caption{Plot of the non-condensed fraction $N_{\mathbf{k}\neq0}/N$ as a function of the hold time $\tau_\text{hold}$.  The blue dashed line is the Bogoliubov integral approximation given in Eq.~(\ref{eqBogFraction1D}), the purple dot-dashed line is the Bogoliubov summation expression given in Eq.~(\ref{eqCondensateDepletion}), and the solid black line is the result of a truncated Wigner simulation.  Parameters: $L=600\,\mu\text{m}$, $N=4 \times 10^5$, $n_a=n_b=2\times 10^5$, $\chi_{aa} = 2.67 \times 10^{-2}\, \text{s}^{-1}$, $\chi_{ab}=\chi_{bb}=0$, $\theta=0$.}
  \label{figBogSqueezingValidation}
\end{figure}

@@

We wish to examine the scaling of the non-condensed fraction $N_{\mathbf{k}\neq0}/N$ with the physical size of the condensate.  Other system parameters such as the total number $N$ and the occupation of the $\ket{a}$ and $\ket{b}$ states at $t_1$ ($n_a$ and $n_b$, respectively) are kept constant and chosen to maximise the squeezing of the equivalent two-mode model presented in Section~\ref{secTwoModeAnalytic}.  In particular this fixes the values of $\theta$, $\phi$ and $\lambda_{ij} = \chi_{ij}\tau_\text{hold}$.  As the size of the condensate is scaled, $\hbar \chi_{ij} = U^\text{1D}_{ij}/L = U^\text{2D}_{ij}/A = U^\text{3D}_{ij}/V$ will change, but we will vary $\tau_\text{hold}$ to keep $\lambda_{ij}$ constant.  Therefore $\Lambda_a = n_a\chi_{aa}\tau_\text{hold}$ will also be constant.  With this in mind, we can write $\tau_\text{hold}$ in terms of the system size and the nonlinearity $U_{aa}$.  The non-condensed fraction can then be shown to scale with the physical size of the condensate as:
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \sqrt{L} \cdot \frac{1}{N} \sqrt{\frac{M U^\text{1D}_{aa} n_a}{2 \hbar^2 \pi^2}} \Lambda_a^{3/2} \cos^2\theta f^\text{1D}(\Lambda_a) &&\propto \sqrt{L}, \\
  \frac{N^\text{2D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx 1 \cdot \frac{1}{N} \frac{M U^\text{2D}_{aa} n_a }{4 \hbar^2} \Lambda_a \cos^2\theta f^\text{2D}(\Lambda_a) &&\propto \text{constant}, \\
  \frac{N^\text{3D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \frac{1}{\sqrt{V}} \cdot \frac{1}{N} \left(\frac{M U^\text{3D}_{aa} n_a}{\hbar^2}\right)^{3/2} \frac{\Lambda_a^{1/2} \cos^2\theta f^\text{3D}(\Lambda_a)}{\sqrt{2} \pi^2} &&\propto \frac{1}{\sqrt{V}}.
\end{align}
The non-condensed fraction $N_{\mathbf{k}\neq 0}/N$ thus scales differently with trap size in each of 1D, 2D and 3D.  For one-dimensional traps, the non-condensed fraction increases with system size, whereas for 2D traps it is constant.  For three-dimensional traps, the non-condensed fraction \emph{decreases} with volume.  This suggests that the best system to use will be a three-dimensional box potential, and the system size can be increased until the condensate depletion becomes negligibly small.

Figures~\ref{fig1D_squeezing_and_bog_scaling}--\ref{fig3D_Bog_mode_occupation_scaling} show the non-condensed fraction predicted by the Bogoliubov analysis compared to that predicted by a full numerical simulation in one, two and three dimensions, as a function of the trap size. Also shown in the 1D and 2D cases is the maximum amount of number squeezing obtainable, as predicted by the numerical simulation. The full squeezing scaling curve is not shown in 3D, as for the smaller trap sizes the TWA method breaks down, and results are unreliable. For the larger trap sizes with small non-condensate fractions, however, the TWA simulation remained valid, and we obtained relative number squeezing of 12.7 dB. The traps are assumed to have a constant potential, which minimises the effects of the multimode issues described in Section \ref{sec:MMdescription}.  We see that the squeezing does indeed degrade at large trap sizes in 1D, but remains constant in 2D. For small trap sizes, the squeezing in 1D asymptotically approaches that of the ideal 0D system.

@@

The simulation shown in Fig.~\ref{fig3D_Bog_mode_occupation_scaling} demonstrates that provided particle number is kept fixed, the number of accessible states for three-dimensional traps decreases with volume in 3D.  This is because the energy of the ground state of the BEC (the chemical potential, $\mu$) is reducing faster than the density of states at low energies is increasing, which in turn results in fewer states that can become dynamically occupied.  This scaling of the estimated Bogoliubov fraction in 3D suggests that strong squeezing will be found in very large, weakly trapped condensates, provided the trapping potentials are box-like. Such box-like traps have already been demonstrated and shown to be capable of producing BECs \cite{gaunt2013}.

@@

As a demonstration of the plausibility of such a scheme, we carried out a full 3D simulation using the geometry of the optical box potential described in Ref.~\cite{gaunt2013}, using the same potentials (a cylindrical ``box'' with length 63\,$\mu$m and diameter 30\,$\mu$m) and condensate number ($N=10^5$ $^{87}$Rb atoms). We set one s-wave scattering length to $5$\,nm and the others to zero: $a_{00}=5$\,nm, $a_{10}=a_{11}=0$. The associated nonlinearities are given by Eq.~(\ref{eqUij}).  This system produced number squeezing of 15dB, and an extremely low non-condensed fraction of $4 \times 10^{-4}$. Both of these numbers are in agreement with the Bogoliubov calculations.

@@

The large Kerr-type nonlinearities present between neutral atoms in BEC present the possibility of significant squeezing, but they also cause potentially complicated multimode behaviour when large BECs are used.  Furthermore, the nonlinearities between multiple internal atomic states are not yet known for many atomic species, and often have sensitive dependence on external magnetic fields due to Feshbach resonances \cite{cornish:2000,altin:2010}.  This paper analyses the requirements for producing useful squeezing via these nonlinearities for systems with large numbers of atoms.

We have presented an analytic solution to the two-mode model of atomic fields with a strong Kerr nonlinearity, allowing for a rapid estimate of the atomic number difference squeezing that can be obtained from these systems as scattering length data is obtained. In general, it predicts arbitrarily good squeezing in the two-mode limit. We then examined trap geometries in which this limit can be approached.

@@

\bibitem{roosET2006} Roos C F, Chwalla M, Kim K, Riebe M, and Blatt R Designer atoms for
quantum metrology 2006 \emph{Nature} {\bf 443}, 316

@@

\bibitem{altinET2013} Altin P A \emph{et al.} Precision atomic gravimeter based on Bragg diffraction 2013 \emph{New Journal of Physics} {\bf 15} 023009
